SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source Term Evaluation

### 2018-02-17 paper.tex
update

\begin{table}[htb]
\centering
\begin{tabular}{@{}l l l l l@{}}
\toprule
CPU Model & Instruction Set & Vector Width & Cores & Identifier \\
\midrule
Xeon X5650 & SSE4.2 & \SI{128}{\bit} & $2 \times 6$ & \texttt{sse4.2} \\
E5-2690 V3 & AVX2 & \SI{256}{\bit} & $2 \times 12$ & \texttt{avx2} \\
\bottomrule
\end{tabular}
\caption{The Intel CPU configurations used in this study. The identifier field will be used as a shorthand descriptor in the performance plots to quickly identify the CPU type.}
\label{t:cpus}
\end{table}

\begin{table}
\centering
\begin{tabular}{@{}l l l l@{}}
\toprule
NVIDIA Model & Driver Version & CUDA Cores & Identifier \\
\midrule
Tesla C2075 & \num{384.81} & \num{448} & \texttt{C2075} \\
Tesla K40m & \num{387.26} & \num{2880} & \texttt{K40m} \\
\bottomrule
\end{tabular}
\caption{The NVIDIA GPU configurations used in this study.  NVIDIA's OpenCL runtime is provided with the graphic driver, rather than any specific version of CUDA.
Again, the identifier field will be used as a shorthand descriptor during analysis of the performance results.}
\label{t:gpus}
\end{table}

@@

However, \texttt{pyJac} allows the user to simply ignore these derivatives (via a command-line switch) to and avoid any negative adverse affects on the Jacobian sparsity.
The rational behind this choice is that many common implicit integration techniques (e.g., CVODE~\cite{cvode:2.8.2}) used in solution of chemical kinetic initial value problems are formulated around the assumption that the supplied Jacobian is approximate; this allows the Jacobian and its LU factorization to be reused for multiple internal integration time-steps, accelerating the solution process.
For such solvers, there is no need for the exact form of the Jacobian and the so-called ``appropriate'' form is preferable.
Hence, in this section we will detail the sparsity of both forms of the Jacobian for the chemical models tested.

@@

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on a variety of different machines.

four ten-core \SI{2.2}{\giga\hertz} Intel Xeon E5-4640 v2 CPUs with \SI{20}{\mega\byte} of L3 cache memory, installed on an Ace Powerworks PW8027R-TRF+ with a Supermicro X9QR7-TF+/X9QRi-F+ baseboard.
Runtimes in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
All code was compiled with \texttt{gcc 4.8.5} and ``\texttt{-O3}'' optimization, running on  of the

\begin{figure}[htb]
  \centering
  \begin{minipage}[t]{0.48\linewidth}
    \includegraphics[width=\textwidth]{SIMD_SIMT_Order_comparison}
    \caption{Runtime in milliseconds per thermo-chemical state of shallow-SIMD and SIMT-vectorization on a single CPU core for both ``C'' (filled symbols) and ``F'' (empty symbols) data ordering.}
    \label{F:SIMDComp}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{0.48\linewidth}
    \includegraphics[width=\textwidth]{SIMD_Vecwidth_comparison}
    \caption{Speedup of a ``C''-ordered shallow-SIMD vectorization on a single CPU core when varying the supplied vector-width.}
    \label{F:Veccomp}
  \end{minipage}
\end{figure}
\begin{figure}[htb]
  \centering
  \begin{minipage}[t]{0.5\linewidth}
    \includegraphics[width=\textwidth]{SIMD_scaling}
    \caption{Strong parallel scaling efficiency of a ``C''-ordered shallow-SIMD vectorization using a vector-width of 16.}
    \label{F:SIMDscale}
  \end{minipage}
\end{figure}

In Fig.~\ref{F:SIMDComp}, the performance of the shallow-SIMD and SIMT vectorization are compared on a single CPU core for both ``C'' and ``F'' memory formats.
The ``C''-ordered SIMD\slash SIMT vectorizations show a speedup over the equivalent ``F''-ordered code by \SIrange{1.46}{2.00}{$\times$} and \SIrange{1.61}{2.38}{$\times$} respectively.
This demonstrates that enhanced data-locality of the ``C''-ordering for CPU caching---discussed in Sec.~\ref{S:data}---is key to high performance on the CPU, and suggests that a deep vectorization would see even greater accelerations.
The shallow vectorization is consistently faster than the SIMT vectorization, by \SIrange{1.73}{1.90}{$\times$} and \SIrange{2.00}{2.45}{$\times$} for the ``C'' and ``F'' memory formats respectively.
This is less than the theoretical maximum speedup on this CPU, which can process \SI{256}{\bit} AVX instructions (i.e. four double-precision operations) concurrently.
Most likely this is due to the non-sequential memory access patterns of the shallow vectorization discussed previously, but could also be caused in part by inefficiencies in translation of the OpenCL code to AVX code, a process handled by the Intel Opencl Runtime.

The effect of changing the vector-width supplied to a ``C''-ordered shallow-SIMD vectorization on a single CPU core is investigated in Fig.~\ref{F:Veccomp}.
Significant speedups are seen with larger vector-widths for the smaller models, however as the model size increases data must be loaded from increasingly distant memory locations and the performance of all three vector-widths becomes similar.

Finally, the strong parallel scaling efficiency is defined as:
\begin{equation}
S = \frac{t_1}{N t_N}
\end{equation}
where $t_1$ is the mean wall-clock runtime of thermo-chemical source terms evaluation on one core and $t_N$ the mean runtime on $N$ cores.
Fig.~\ref{F:SIMDscale} examines the strong parallel scaling efficiency for all four models with a ``C''-ordered shallow SIMD-vectorization and vector-width of 16.
Interestingly the efficiency of the hydrogen model decreases the fastest while the larger models have significantly larger scaling efficiencies for more than 4 cores.
Additionally, it appears that for a fixed number of cores, the scaling efficiency increases with model size.
This suggests that evaluation of the chemical-source terms for a single thermo-chemical state for the smaller models may not saturate the processing power of the CPU, some efficiency may be lost due to context switching within OpenCL itself.
Additionally, OpenCL allows assigning of work to devices by shared memory resources, e.g., a L3 cache, which was not pursued in this study for simplicity; this should be investigated in future work to determine the effect on scaling efficiency.

>>>>

\begin{table}[htb]
\centering
\begin{tabular}{@{}l l l@{}}
\toprule
CPU Model        & Xeon X5650      & E5-2690 V3     \\
\midrule
Instruction Set  & SSE4.2 	   & AVX2 	    \\
Vector Width     & \SI{128}{\bit}  & \SI{256}{\bit} \\
Cores            & $2 \times 6$    & $2 \times 12$  \\
Identifier       & \texttt{sse4.2} & \texttt{avx2}  \\
OpenCL Version   & \num{1.2}       & \num{1.2}      \\
\bottomrule
\end{tabular}
\caption{The Intel CPU configurations used in this study. The identifier field will be used as a shorthand descriptor in the performance plots to quickly identify the CPU type.}
\label{t:cpus}
\end{table}

\begin{table}
\centering
\begin{tabular}{@{}l l l l l@{}}
\toprule
NVIDIA Model   & Tesla C2075    & Tesla K40m    \\
\midrule
Driver Version & \num{384.81}   & \num{387.26}  \\
CUDA Cores     & \num{448}      & \num{2880}    \\
Identifier     & \texttt{C2075} & \texttt{K40m} \\
OpenCL Version & \num{1.1}	& \num{1.2}	\\
\bottomrule
\end{tabular}
\caption{The NVIDIA GPU configurations used in this study.  NVIDIA's OpenCL runtime is provided with the graphic driver, rather than any specific version of CUDA.
Again, the identifier field will be used as a shorthand descriptor during analysis of the performance results.}
\label{t:gpus}
\end{table}

@@

However, \texttt{pyJac} allows the user to simply ignore these derivatives (via a command-line switch) to and avoid any negative adverse affects on the Jacobian sparsity.
The rational behind this choice is that many common implicit integration techniques (e.g., CVODE~\cite{cvode:2.8.2}) used in solution of chemical kinetic initial value problems are formulated around the assumption that the supplied Jacobian is approximate; this allows the Jacobian and its LU factorization to be reused for multiple internal integration time-steps, accelerating the solution process.
For such solvers, there is no need for the exact form of the Jacobian and the so-called ``approximate'' form is preferable.
Hence, in this section we will detail the sparsity of both forms of the Jacobian for the chemical models tested.

@@

\subsection{Performance}
\label{S:results}
The performance studies in this work were run on the platforms listed in~\cref{t:cpus,t:gpus}.
Runtimes in each case were averaged over ten runs, each using the same set of PaSR conditions utilized in validation.
The wrapping code---responsible for initializing\slash transfering memory, reading input, etc.---was compiled with \texttt{gcc v5.4.0} on the \texttt{avx2}\slash\texttt{K40m} platforms, and \texttt{gcc 4.8.5} on the \texttt{sse4.2}\slash\texttt{C2075} machines.
The optimization level ``\texttt{-O3}'' was used in all cases, and no ``fast-math'' OpenCL optimizations were enabled.
Additionally, the ``exact'' form of the Jacobian (as opposed the the ``appropriate'' form discussed in~\cref{S:sparsity}) was used in all cases.

\Cref{f:source} shows the runtime of \texttt{pyJac} on the test platforms.

\begin{figure}[htb]
   \centering
  \begin{subfigure}{0.45\linewidth}
      \includegraphics[width=\textwidth]{intel_source.pdf}
      \caption{The ``exact'' Jacobian.}
      \label{F:intel_source_noorm}
  \end{subfigure}
 \label{f:source}
\end{figure}
